---
title: "Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

library(basetable)
library(bench)
library(dplyr)
library(ggplot2)
library(data.table)
```

The benchmarks below compare `basetable` with raw `data.table`, base R, and
`dplyr` on four representative table operations:

- row subsetting and column selection
- keyed joins
- grouped aggregation
- grouped counting

`basetable` wraps `data.table` as its execution backend, so the `data.table`
comparison is the one that matters most here: it isolates the cost of the
base-R-faithful wrapper itself, separate from the backend's own performance.
The goal is not to prove a universal winner over base R or dplyr. It is to
show how close the wrapper gets to its own backend.

```{r data}
set.seed(1)

subset_n <- 1e5
merge_n <- 5e4

subset_dt <- data.table(
  id = seq_len(subset_n),
  grp = sample(letters[1:5], subset_n, replace = TRUE),
  value = rnorm(subset_n)
)

merge_x_dt <- data.table(
  id = seq_len(merge_n),
  value_x = rnorm(merge_n)
)

merge_y_dt <- data.table(
  id = sample(seq_len(merge_n), merge_n),
  value_y = rnorm(merge_n)
)
```

```{r benchmark}
run_subset_benchmark <- function() {
  bench::mark(
    basetable = basetable::subset(
      subset_dt,
      value > 0,
      select = c("id", "grp", "value")
    ),
    data_table = subset_dt[value > 0, .(id, grp, value)],
    base_r = base::subset(
      subset_dt,
      value > 0,
      select = c("id", "grp", "value")
    ),
    dplyr = subset_dt |>
      dplyr::filter(value > 0) |>
      dplyr::select(id, grp, value),
    iterations = 15,
    check = FALSE
  )
}

run_merge_benchmark <- function() {
  bench::mark(
    basetable = basetable::merge(merge_x_dt, merge_y_dt, by = "id"),
    data_table = data.table::merge.data.table(merge_x_dt, merge_y_dt, by = "id"),
    base_r = base::merge(merge_x_dt, merge_y_dt, by = "id"),
    dplyr = dplyr::inner_join(merge_x_dt, merge_y_dt, by = "id"),
    iterations = 15,
    check = FALSE
  )
}

run_aggregate_benchmark <- function() {
  bench::mark(
    basetable = basetable::aggregate(
      subset_dt,
      by = "grp",
      value = "value",
      fun = mean,
      sort = FALSE
    ),
    data_table = subset_dt[, .(value = mean(value)), by = grp],
    base_r = stats::aggregate(
      value ~ grp,
      data = subset_dt,
      FUN = mean
    ),
    dplyr = subset_dt |>
      dplyr::group_by(grp) |>
      dplyr::summarise(value = mean(value), .groups = "drop"),
    iterations = 15,
    check = FALSE
  )
}

run_count_benchmark <- function() {
  bench::mark(
    basetable = basetable::count(subset_dt, by = "grp"),
    data_table = subset_dt[, .N, by = grp],
    base_r = as.data.frame(table(subset_dt$grp)),
    dplyr = subset_dt |> dplyr::count(grp),
    iterations = 15,
    check = FALSE
  )
}

summarise_benchmark <- function(x, operation) {
  out <- data.table::as.data.table(x)[, .(
    operation = operation,
    implementation = as.character(expression),
    median_ms = as.numeric(median) * 1000,
    itr_per_sec = as.numeric(`itr/sec`),
    mem_alloc_mb = as.numeric(mem_alloc) / 1024^2
  )]

  out[, relative_time := median_ms / median_ms[implementation == "basetable"], by = operation]
  out[, overhead_vs_data_table := median_ms / median_ms[implementation == "data_table"], by = operation]
  out[, implementation := factor(
    implementation,
    levels = c("basetable", "data_table", "base_r", "dplyr"),
    labels = c("basetable", "data.table", "base R", "dplyr")
  )]
  out
}

benchmark_results <- data.table::rbindlist(list(
  summarise_benchmark(run_subset_benchmark(), "Subset and select"),
  summarise_benchmark(run_merge_benchmark(), "Merge"),
  summarise_benchmark(run_aggregate_benchmark(), "Aggregate"),
  summarise_benchmark(run_count_benchmark(), "Group count")
), fill = TRUE)

benchmark_results[, operation := factor(operation, levels = c("Subset and select", "Merge", "Aggregate", "Group count"))]
```

## Results Table

```{r table, results='asis'}
table_results <- data.table::copy(benchmark_results)
data.table::setorderv(table_results, c("operation", "implementation"))
table_results[, `:=`(
  median_ms = format(round(median_ms, 2), nsmall = 2),
  itr_per_sec = format(round(itr_per_sec, 1), nsmall = 1),
  mem_alloc_mb = format(round(mem_alloc_mb, 2), nsmall = 2),
  relative_time = format(round(relative_time, 2), nsmall = 2),
  overhead_vs_data_table = format(round(overhead_vs_data_table, 2), nsmall = 2)
)]

knitr::kable(
  table_results,
  col.names = c(
    "Operation",
    "Implementation",
    "Median (ms)",
    "Iterations / sec",
    "Memory (MB)",
    "Relative time vs basetable",
    "Relative time vs data.table"
  ),
  align = c("l", "l", "r", "r", "r", "r", "r")
)
```

The "Relative time vs data.table" column is the direct measure of wrapper
cost: a value close to 1.00 for the `basetable` row means the base-R-faithful
interface adds effectively no overhead over calling `data.table` directly.

A broader perf pass also rewrote several other functions that previously did
their grouping/joining/row iteration in pure base R (`interaction()`+`table()`
or `split()`, row-wise `apply()`, `paste()`-based key matching) instead of
using data.table's compiled internals. `inst/benchmarks/benchmark-all.R`
tracks these against their native data.table equivalents; the most dramatic
fix was `rollingmerge()`, which reimplemented a rolling/nearest join with a
hand-rolled per-group R loop and measured at roughly **1500x** the cost of
data.table's native `roll=` join before being rewritten to use it directly —
it now runs at parity. `count()`, `duplicated_keys()`, `freq()`, `filldown()`,
`fillup()`, `split()`, and `summarise()`/`summarize()`/`summaries()` all saw
similar (smaller) reductions in overhead from the same kind of fix.

## Absolute Time

```{r absolute-plot, fig.cap="Median runtime by implementation and operation."}
ggplot(
  benchmark_results,
  aes(x = implementation, y = median_ms, fill = implementation)
) +
  geom_col(width = 0.7) +
  facet_wrap(~operation, ncol = 1, scales = "free_y") +
  coord_flip() +
  scale_y_log10(labels = scales::label_number(accuracy = 0.01)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = NULL,
    y = "Median time (ms, log scale)",
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
```

## Relative Time

```{r relative-plot, fig.cap="Relative runtime compared with basetable for each workload. Values above 1 mean slower than basetable."}
ggplot(
  benchmark_results,
  aes(x = implementation, y = relative_time, fill = implementation)
) +
  geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
  geom_col(width = 0.7) +
  facet_wrap(~operation, ncol = 1, scales = "free_y") +
  coord_flip() +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = NULL,
    y = "Relative time vs basetable",
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
```
