---
title: "Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 6
)
library(basetable)
library(bench)
library(ggplot2)

have <- function(pkg) requireNamespace(pkg, quietly = TRUE)
opt <- c(data.table = have("data.table"), dplyr = have("dplyr"), collapse = have("collapse"))

# Keep the vignette build quick; the standalone harness in
# inst/benchmarks/benchmark-scale.R runs the full matrix.
N     <- as.integer(Sys.getenv("BT_VIGNETTE_N", "3e5"))
REPS  <- as.integer(Sys.getenv("BT_VIGNETTE_REPS", "7"))
basetable::setthreads(percent = 100)
```

`basetable` runs every operation on its own C++ engine and has **no
`data.table` dependency**. This vignette compares it with **base R**,
**`data.table`**, **`dplyr`** and **`collapse`** (each optional, skipped if
absent) across the operations where the differences are most informative.

```{r data}
set.seed(1)
n <- N
d <- data.frame(
  g  = sprintf("k%07d", sample(2000L, n, replace = TRUE)),        # ~2000 groups
  gh = sprintf("k%08d", sample(max(n %/% 10L, 1L), n, replace = TRUE)),  # ~n/10 groups
  x  = rnorm(n),
  y  = rnorm(n),
  id = seq_len(n),
  stringsAsFactors = FALSE
)
dim_tbl <- d[!duplicated(d$g), c("g", "y")]

if (opt[["data.table"]]) {
  dt  <- data.table::as.data.table(d)
  dmt <- data.table::as.data.table(dim_tbl)
}
```

```{r run}
bench_one <- function(label, exprs) {
  exprs <- exprs[vapply(names(exprs), function(nm)
    nm %in% c("basetable", "base R") || isTRUE(opt[[nm]]), logical(1))]
  m <- bench::mark(exprs = exprs, iterations = REPS, check = FALSE, memory = TRUE)
  data.frame(
    operation      = label,
    implementation = as.character(m$expression),
    median_ms      = as.numeric(m$median) * 1000,
    mem_mb         = as.numeric(m$mem_alloc) / 1024^2,
    stringsAsFactors = FALSE
  )
}

results <- rbind(
  bench_one("filter", list(
    basetable  = quote(basetable::subset(d, x > 0.5)),
    `base R`   = quote(d[d$x > 0.5, , drop = FALSE]),
    data.table = quote(dt[x > 0.5]),
    dplyr      = quote(dplyr::filter(d, x > 0.5)),
    collapse   = quote(collapse::fsubset(d, x > 0.5)))),
  bench_one("sort (string key)", list(
    basetable  = quote(basetable::orderrows(d, by = c("g", "x"))),
    `base R`   = quote(d[order(d$g, d$x), , drop = FALSE]),
    data.table = quote(data.table::setorder(data.table::copy(dt), g, x)),
    dplyr      = quote(dplyr::arrange(d, g, x)),
    collapse   = quote(collapse::roworder(d, g, x)))),
  bench_one("distinct", list(
    basetable  = quote(basetable::uniquerows(d, cols = "g")),
    `base R`   = quote(unique(d[, "g", drop = FALSE])),
    data.table = quote(unique(dt[, list(g)])),
    dplyr      = quote(dplyr::distinct(d, g)),
    collapse   = quote(collapse::funique(d["g"])))),
  bench_one("count by group", list(
    basetable  = quote(basetable::count(d, by = "gh", sort = FALSE)),
    `base R`   = quote(as.data.frame(table(d$gh))),
    data.table = quote(dt[, .N, by = gh]),
    dplyr      = quote(dplyr::count(d, gh)),
    collapse   = quote(collapse::fcount(d, gh)))),
  bench_one("sd by group", list(
    basetable  = quote(basetable::aggregate(d, by = "g", value = "x", fun = sd, sort = FALSE)),
    `base R`   = quote(tapply(d$x, d$g, sd)),
    data.table = quote(dt[, list(x = sd(x)), by = g]),
    dplyr      = quote(dplyr::summarise(dplyr::group_by(d, g), x = sd(x), .groups = "drop")),
    collapse   = quote(collapse::fsd(d$x, d$g)))),
  bench_one("equi join", list(
    basetable  = quote(basetable::merge(d, dim_tbl, by = "g")),
    `base R`   = quote(merge(d, dim_tbl, by = "g")),
    data.table = quote(merge(dt, dmt, by = "g")),
    dplyr      = quote(dplyr::inner_join(d, dim_tbl, by = "g")),
    collapse   = quote(collapse::join(d, dim_tbl, on = "g", how = "inner", verbose = 0)))),
  bench_one("semi join", list(
    basetable  = quote(basetable::semimerge(d, dim_tbl, by = "g")),
    `base R`   = quote(d[d$g %in% dim_tbl$g, , drop = FALSE]),
    data.table = quote(dt[dmt, on = "g", nomatch = NULL]),
    dplyr      = quote(dplyr::semi_join(d, dim_tbl, by = "g")),
    collapse   = quote(collapse::fsubset(d, g %in% dim_tbl$g))))
)

order_ops <- c("filter", "sort (string key)", "distinct", "count by group",
               "sd by group", "equi join", "semi join")
results$operation <- factor(results$operation, levels = order_ops)
bt <- results[results$implementation == "basetable", c("operation", "median_ms")]
names(bt)[2] <- "bt_ms"
results <- merge(results, bt, by = "operation")
results$vs_basetable <- results$median_ms / results$bt_ms
```

## Results

```{r table, results='asis'}
tab <- results[order(results$operation, results$implementation),
               c("operation", "implementation", "median_ms", "mem_mb", "vs_basetable")]
for (col in c("median_ms", "mem_mb", "vs_basetable"))
  tab[[col]] <- format(round(tab[[col]], 2), nsmall = 2)
knitr::kable(
  tab, row.names = FALSE,
  col.names = c("Operation", "Implementation", "Median (ms)", "Memory (MB)",
                "Relative to basetable"),
  align = c("l", "l", "r", "r", "r")
)
```

## Time relative to basetable

```{r plot, fig.cap="Median runtime by engine, relative to basetable (dashed line). Below 1 is faster than basetable."}
ggplot(results, aes(implementation, vs_basetable, fill = implementation)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey50") +
  geom_col(width = 0.7) +
  facet_wrap(~operation, ncol = 2, scales = "free_x") +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Median time / basetable", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))
```

## Reading the results

`basetable` beats `data.table` on `distinct` and grouped `count`, is close on
`sd`-by-group and `merge`, and trails on `filter` and `sort`. Against base R
and `dplyr` it is often an order of magnitude faster on the grouped
operations. It also allocates far less: a grouped `aggregate` or `count`
touches near-zero memory because the reducers accumulate in C++ without
materialising intermediate columns.

The gap grows in `basetable`'s favour at larger sizes and higher
cardinality: `benchmark-scale.R` at 1e6 rows with ~100k groups shows
`semimerge` at ~0.25x of `data.table` and `count` / `sd` well below 1. The
one operation it does not match at any size is **sorting** -- `orderrows()`
uses a stable parallel radix over integer codes and is ~20x faster than base
`order()`, but `data.table`'s hand-tuned parallel radix stays ahead.

Numbers are machine- and size-specific. Set `BT_VIGNETTE_N` and
`BT_VIGNETTE_REPS`, or run `inst/benchmarks/benchmark-scale.R`, to reproduce
at other sizes and cardinalities.
