---
title: "Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

library(basetable)
library(bench)
library(ggplot2)

have <- function(pkg) requireNamespace(pkg, quietly = TRUE)
opt <- c(
  data.table = have("data.table"),
  dplyr = have("dplyr"),
  collapse = have("collapse"),
  polars = have("polars")
)
```

`basetable` runs every table operation through its own native C++ engine and
has **no `data.table` dependency**. These benchmarks still line it up against
the fastest data-frame tools in R -- **base R**, **`data.table`**, **`dplyr`**,
**`collapse`**, and **`polars`** -- on four representative operations:

- row subsetting and column selection
- keyed joins
- grouped aggregation
- grouped counting

Arms whose package is not installed are dropped automatically. The point is
not to crown a winner but to show that a base-R-style interface backed by
compiled kernels stays competitive with the established engines.

```{r data}
set.seed(1)

subset_n <- 1e5
merge_n <- 5e4

tbl <- data.frame(
  id = seq_len(subset_n),
  grp = sample(letters[1:5], subset_n, replace = TRUE),
  value = rnorm(subset_n),
  stringsAsFactors = FALSE
)
merge_x <- data.frame(id = seq_len(merge_n), value_x = rnorm(merge_n))
merge_y <- data.frame(id = sample(seq_len(merge_n), merge_n), value_y = rnorm(merge_n))

if (opt[["data.table"]]) {
  tbl_dt <- data.table::as.data.table(tbl)
  mx_dt <- data.table::as.data.table(merge_x)
  my_dt <- data.table::as.data.table(merge_y)
}
if (opt[["polars"]]) {
  tbl_pl <- polars::as_polars_df(tbl)
  mx_pl <- polars::as_polars_df(merge_x)
  my_pl <- polars::as_polars_df(merge_y)
}
```

```{r benchmark}
mark_arms <- function(arms) {
  keep <- vapply(names(arms), function(nm) {
    nm %in% c("basetable", "base R") || isTRUE(opt[[nm]])
  }, logical(1))
  bench::mark(exprs = arms[keep], iterations = 15, check = FALSE)
}

run_subset <- function() mark_arms(list(
  basetable  = quote(basetable::subset(tbl, value > 0, select = c("id", "grp", "value"))),
  `base R`   = quote(base::subset(tbl, value > 0, select = c("id", "grp", "value"))),
  data.table = quote(tbl_dt[value > 0, list(id, grp, value)]),
  dplyr      = quote(dplyr::select(dplyr::filter(tbl, value > 0), id, grp, value)),
  collapse   = quote(collapse::fsubset(tbl, value > 0, id, grp, value)),
  polars     = quote(tbl_pl$filter(polars::pl$col("value") > 0)$select("id", "grp", "value"))
))

run_merge <- function() mark_arms(list(
  basetable  = quote(basetable::merge(merge_x, merge_y, by = "id")),
  `base R`   = quote(base::merge(merge_x, merge_y, by = "id")),
  data.table = quote(merge(mx_dt, my_dt, by = "id")),
  dplyr      = quote(dplyr::inner_join(merge_x, merge_y, by = "id")),
  collapse   = quote(collapse::join(merge_x, merge_y, on = "id", how = "inner", verbose = 0)),
  polars     = quote(mx_pl$join(my_pl, on = "id", how = "inner"))
))

run_aggregate <- function() mark_arms(list(
  basetable  = quote(basetable::aggregate(tbl, by = "grp", value = "value", fun = mean, sort = FALSE)),
  `base R`   = quote(stats::aggregate(value ~ grp, data = tbl, FUN = mean)),
  data.table = quote(tbl_dt[, list(value = mean(value)), by = grp]),
  dplyr      = quote(dplyr::summarise(dplyr::group_by(tbl, grp), value = mean(value), .groups = "drop")),
  collapse   = quote(collapse::fsummarise(collapse::fgroup_by(tbl, grp), value = collapse::fmean(value))),
  polars     = quote(tbl_pl$group_by("grp")$agg(polars::pl$col("value")$mean()))
))

run_count <- function() mark_arms(list(
  basetable  = quote(basetable::count(tbl, by = "grp")),
  `base R`   = quote(as.data.frame(table(tbl$grp))),
  data.table = quote(tbl_dt[, .N, by = grp]),
  dplyr      = quote(dplyr::count(tbl, grp)),
  collapse   = quote(collapse::fcount(tbl, grp)),
  polars     = quote(tbl_pl$group_by("grp")$agg(polars::pl$len()))
))

summarise_benchmark <- function(x, operation) {
  out <- data.frame(
    operation = operation,
    implementation = as.character(x$expression),
    median_ms = as.numeric(x$median) * 1000,
    itr_per_sec = as.numeric(x$`itr/sec`),
    mem_alloc_mb = as.numeric(x$mem_alloc) / 1024^2,
    stringsAsFactors = FALSE
  )
  base_ms <- out$median_ms[out$implementation == "basetable"]
  out$relative_time <- out$median_ms / base_ms
  out
}

benchmark_results <- do.call(rbind, list(
  summarise_benchmark(run_subset(), "Subset and select"),
  summarise_benchmark(run_merge(), "Merge"),
  summarise_benchmark(run_aggregate(), "Aggregate"),
  summarise_benchmark(run_count(), "Group count")
))
benchmark_results$operation <- factor(
  benchmark_results$operation,
  levels = c("Subset and select", "Merge", "Aggregate", "Group count")
)
```

## Results table

```{r table, results='asis'}
tab <- benchmark_results[order(benchmark_results$operation, benchmark_results$implementation), ]
for (col in c("median_ms", "itr_per_sec", "mem_alloc_mb", "relative_time")) {
  tab[[col]] <- format(round(tab[[col]], 2), nsmall = 2)
}
knitr::kable(
  tab,
  row.names = FALSE,
  col.names = c("Operation", "Implementation", "Median (ms)",
                "Iterations / sec", "Memory (MB)", "Relative time vs basetable"),
  align = c("l", "l", "r", "r", "r", "r")
)
```

A value near 1.00 in the "Relative time vs basetable" column means the other
implementation runs at roughly the same speed as basetable for that workload.
The standalone `inst/benchmarks/benchmark-competitors.R` harness runs the same
comparison at larger sizes.

## Absolute time

```{r absolute-plot, fig.cap="Median runtime by implementation and operation."}
ggplot(benchmark_results, aes(x = implementation, y = median_ms, fill = implementation)) +
  geom_col(width = 0.7) +
  facet_wrap(~operation, ncol = 1, scales = "free_y") +
  coord_flip() +
  scale_y_log10(labels = scales::label_number(accuracy = 0.01)) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Median time (ms, log scale)", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))
```

## Relative time

```{r relative-plot, fig.cap="Relative runtime compared with basetable. Values above 1 mean slower than basetable."}
ggplot(benchmark_results, aes(x = implementation, y = relative_time, fill = implementation)) +
  geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
  geom_col(width = 0.7) +
  facet_wrap(~operation, ncol = 1, scales = "free_y") +
  coord_flip() +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Relative time vs basetable", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))
```
