---
title: "Data manipulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data manipulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(basetable)
```

`basetable` uses compact verbs with explicit arguments. The same functions
work in nested expressions and native R pipelines, always returning a
`data.table` without modifying the input.

## A compact pipeline

```{r}
mtcars |>
  subset(cyl >= 6, select = c("mpg", "hp", "wt", "cyl")) |>
  transform(power = hp / wt) |>
  reorder(by = c("cyl", "mpg"), decreasing = c(FALSE, TRUE))
```

`reorder()` and `orderrows()` accept one direction per column. This makes a
mixed ascending/descending order explicit without a descending-expression
mini-language.

```{r}
orderrows(
  mtcars,
  by = c("cyl", "mpg"),
  decreasing = c(FALSE, TRUE)
) |>
  firstrows(6)
```

## Values from the calling function

Transformation expressions can combine table columns with ordinary values
defined by the calling function. Newly created columns are available to later
expressions in the same call.

```{r}
scorecars <- function(data, horsepowerweight = 0.7) {
  data |>
    mutate(
      weightedhp = hp * horsepowerweight,
      score = weightedhp / wt
    ) |>
    orderrows("score", decreasing = TRUE)
}

scorecars(mtcars) |>
  pick(c("mpg", "hp", "wt", "score")) |>
  firstrows(5)
```

## Grouped summaries

```{r}
aggregate(airquality, by = "Month", value = c("Ozone", "Temp"), fun = mean, na.rm = TRUE)
```

The expression-oriented equivalent stays compact when several summaries use
different functions.

```{r}
summarise(
  airquality,
  ozone = mean(Ozone, na.rm = TRUE),
  temperature = mean(Temp, na.rm = TRUE),
  days = length(Temp),
  by = "Month"
)
```

## Joins

```{r}
merge(
  data.frame(id = 1:3, x = letters[1:3]),
  data.frame(id = c(2, 3, 4), y = LETTERS[2:4]),
  by = "id",
  all = TRUE
)
```

## Validation in a pipeline

Assertions return the input invisibly when they pass, so a pipeline can fail
close to the operation that violated its contract.

```{r}
cars <- mtcars |>
  mutate(car = rownames(mtcars)) |>
  firstcols("car")

assertcomplete(cars, c("car", "mpg", "cyl"))
assertunique(cars, "car")
assertrows(cars, mpg > 0)
```

## Explicit package calls

Several compact names intentionally match base R or other table packages.
In scripts that attach multiple table packages, qualify the verb rather than
changing the workflow grammar.

```{r}
mtcars |>
  basetable::filter(cyl == 6) |>
  basetable::pick(c("mpg", "hp", "wt")) |>
  basetable::mutate(power = hp / wt)
```
