---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(basetable)
```

`basetable` provides a base-style front end for tabular work with a
`data.table` backend. The core design goals are explicit semantics, compact
naming, and equal support for nested calls and native `|>` pipelines.

## The canonical compact core

Most workflows need only a small set of regular verbs:

| Task | Canonical functions |
|:--|:--|
| Inspect | `glimpse()`, `describe()`, `missingness()` |
| Keep rows | `subset()` |
| Keep or remove columns | `pick()`, `drop()` |
| Create columns | `transform()` |
| Order rows | `reorder()`, `orderrows()` |
| Summarise | `aggregate()`, `summaries()`, `count()` |
| Join | `merge()`, `antimerge()`, `semimerge()` |
| Reshape | `tolong()`, `towide()` |
| Validate | `assertcomplete()`, `assertunique()`, `assertrows()` |

These are canonical names, not one layer of aliases among several grammars.
Each function accepts a table as its first argument, uses character vectors
for column specifications, and returns a `data.table`. Those regular rules
make pipelines compact for people and make function selection predictable for
program-generating tools such as language models.

The package is intentionally focused on in-memory table manipulation,
exploration, and validation. Read and write files with base R or a dedicated
I/O package, then pass the resulting data frame or data.table to basetable.
Domain-specific parsing and validation remain the responsibility of focused
packages.

```{r}
subset(mtcars, cyl == 6, select = c("mpg", "hp", "wt"))
```

```{r}
mtcars |>
  pick(c("mpg", "hp", "wt", "cyl")) |>
  aggregate(by = "cyl", value = c("mpg", "hp"), fun = mean)
```

## Namespaces stay explicit

Compact names intentionally overlap with base R and other table packages. If
more than one is attached, qualify the function rather than introducing a new
alias or changing grammars mid-pipeline.

```{r}
mtcars |>
  basetable::subset(cyl == 6) |>
  basetable::pick(c("mpg", "hp", "wt")) |>
  basetable::transform(power = hp / wt)
```
